Consistent ecosystem functional response during five years of drought and deluge in a sagebrush steppe

### 2017-07-11 manuscript/usses_water_ms.Rmd
trivia
more intro writing

---
title: Consistent ecosystem functional response during five years of drought and deluge in a sagebrush steppe
preprint: true
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California xxxxx
abstract: >
  Precipitation is predicted to become more variable in the western U.S., meaning years of above and below average precipitation will become more common. Such periods of drought and deluge could become major drivers of plant community dynamics and ecosystem functioning in water limited grasslands. Here we report the results of a five-year experiment where we used drought and irrigation treatments (50% decrease/increase) to see how a sagebrush steppe plant community in Idaho will respond to future climate changes. The ecosystem was suprisingly resistant to both drought and irrigation. Aboveground net primary productivity (ANPP) responded positively to ambient growing season precipitation, but the response did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functioning was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functioning and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation.
header-includes: >
  \usepackage{lipsum}
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

# Introduction

@@

Here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (50\% decrease/increase) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between ANPP and available soil moisture, and if community dynamics underly the ecosystem responses.
In particular, we are interested in the consistency of the ANPP-soil moisture relationship among treatments.
Is the relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high soil moisture?
To answer these questions we fit a random intercepts, random slopes model to test whether the regressions differed among treatments.

<!---
As the rate of climate change accelerates, ecologists are being pressed to forecast its impacts on ecosystem functions and services.
Making such forecasts faces two challenges.
First, future conditions are likely to be outside the historical range of variability [@Smith2011], meaning we cannot simply look to the past to predict the future.
Second, ecosystems will likely exhibit unique responses to climate change induced resource alterations [e.g., @Byrne2017], meaning we cannot simply look across space to predict the future.
These two challenges motivate the use of *in situ* experimental manipulations of resource availability [@Avolio2015].

@@

# Materials and Methods

@@

To see if community composition differed among treatments through time, we used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis distances.
For each year of the experiment, we first calculated Bray-Curtis distances among all plots, and then extracted those distances for use in the NMDS.
Because we standardized species' abundances, some values were negative, which is not allowed for calculating Bray-Curtis distances.
We simply added '2' to each abundance value to ensure all values were greater than zero.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used the 'metaMDS()' function in the R package 'vegan' [@Oksanen2016] to calculate Bray-Curtis distances and then to run the NMDS analysis.
We used the 'vegan::adonis()' function [@Oksanen2016] to perform permutational multivariate analysis of variance to test whether treatment plots formed distinct groupings.
To test whether treatment plots were equally dispersed, or not, we used the 'vegan::betadisper' function [@Oksanen2016].
Lastly, we examined rank clocks of species' abundances through time to assess the stability of community composition over the course of the experiment [@Collins2008].
Rank clocks were made using 'ggplot2' [@Wickham2009] and R code from the 'codyn' package [@Hallett2016].

@@

# Results

@@

# Discussion

@@

# Acknowledgments

@@

# Funding

@@

# Tables

@@

# Figures

@@

# References

>>>>

---
title: Consistent ecosystem functional response during five years of drought and deluge in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California xxxxx
abstract: >
  Precipitation is predicted to become more variable in the western U.S., meaning years of above and below average precipitation will become more common. Such periods of drought and deluge could become major drivers of plant community dynamics and ecosystem functioning in water limited grasslands. Here we report the results of a five-year experiment where we used drought and irrigation treatments (50% decrease/increase) to see how a sagebrush steppe plant community in Idaho will respond to future climate changes. The ecosystem was suprisingly resistant to both drought and irrigation. Aboveground net primary productivity (ANPP) responded positively to ambient growing season precipitation, but the response did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functioning was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functioning and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation.
header-includes: >
  \usepackage{lipsum}
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

# INTRODUCTION

@@

Another problem with relying on historical ecoystem functional responses is that they are not static.
Changes in species identities and abundances can alter an ecosystem's functional response to water availability because different species have different physiological thresholds for producting biomass.
@Smith2009 introduced the 'Hierarchical Response Framework' (HRF) for understanding the interplay of community composition and ecosystem functioning in response to resource manipulations over time.
In the near term, ecosystem functioning such as annual net primary productivity (ANPP) will reflect the physiological responses of individual species to the manipulated resource level.
For example, ANPP may decline under simulated drought because the initial community consisted of drought-intolerant species [@Hoover2014].
Over longer time spans, ecosystem functioning may recover as new species colonize or initial species reorder in relative abundance.
For example, ANPP may initially decline, but eventually rise back to pre-treatment levels once drought-tolerant species become more abundant and compensate for drought-intolerant species [@Hoover2014].
It is also possible that ecosystem functioning shifts to a new mean state, reflecting the suite of species in the new community [@Knapp2012].

Manipulating potentially limiting resources offers a route to understanding how ecosystems will respond to resource levels that fall outside the historical range of variability [@Avolio2015; @Gherardi2015; @Knapp2017].
Altering the amount of precipitation over many years should provide insight into the time scales at which water-limited ecosystems respond to chronic resource alteration.
Following the HRF, we propose four alternative predictions for the effect of precipitation manipulation on the ecosystem functional response to soil moisture, that is, the soil moisture-ANPP relationship (Figure 1).
The four predictions are based on possible outcomes at the community (e.g., community composition) and ecosystem (e.g., soil moisture-ANPP regression) levels.

First, altered precipitation changes neither ecosystem functional response nor community composition (Fig. 1, top left).
In this case, changes in ANPP simply follow the soil moisture-ANPP relationship under ambient conditions.
This corresponds to the early phases of the HRF, where ecosystem response is due to the physiological responses of individual species.
Second, the ecosystem functional response changes but community composition remains the same (Fig. 1, top right).
A saturating soil moisture-ANPP response fits this scenario, where individual species hit physiological thresholds or are limited by some other resource.
Third, the ecosystem functional response is consistent but underlying community composition changes (Fig. 1, bottom left).
In this case, changes in species' indentities or abundances occur in response to altered precipitation levels and species more suited to the new conditions compensate for reduced function of initial species.
Fourth, and last, both ecosystem functional response and community composition change (Fig. 1, bottom right).
New species, or newly abundant species, with different physiological responses completely reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time scales at which the different scenarios play out likely differ [@Smith2009; @Wilcox2016; @Knapp2017].
Thus, our task is not to test the validity of the HRF, but rather to amass information on how quickly ecosystem functional responses change in different ecosystems.
Likewise, we need to understand whether changes at the ecosystem level are driven community level changes or individual level responses.

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (50\% decrease/increase) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between available soil moisture and ANPP, and if community dynamics underly the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high soil moisture?
To answer these questions we fit a random intercepts, random slopes model to test whether the regressions differed among treatments.
We also analyzed community composition over time, allowing us to place our experimental results within the framework of the HRF and our competing predictions (Fig. 1).

<!---
Chronic alterations to resource availability should cause community composition to shift over time, if the focal ecosystem is sensitive to the manipulated resource.
Species' relative abundances are expected to re-order and, eventually, some species will be lost from the local community and new species will be gained [@Smith2009; @Avolio2015].
On the heels of changes in community composition, subsequent changes in the level of ecosystem functioning may occur.

As the rate of climate change accelerates, ecologists are being pressed to forecast its impacts on ecosystem functions and services.
Making such forecasts faces two challenges.
First, future conditions are likely to be outside the historical range of variability [@Smith2011], meaning we cannot simply look to the past to predict the future.
Second, ecosystems will likely exhibit unique responses to climate change induced resource alterations [e.g., @Byrne2017], meaning we cannot simply look across space to predict the future.
These two challenges motivate the use of *in situ* experimental manipulations of resource availability [@Avolio2015].

@@

# METHODS

@@

To see if community composition differed among treatments through time, we used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis distances.
For each year of the experiment, we first calculated Bray-Curtis distances among all plots, and then extracted those distances for use in the NMDS.
Because we standardized species' abundances, some values were negative, which is not allowed for calculating Bray-Curtis distances.
We simply added '2' to each abundance value to ensure all values were greater than zero.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used the 'metaMDS()' function in the R package 'vegan' [@Oksanen2016] to calculate Bray-Curtis distances and then to run the NMDS analysis.
We used the 'vegan::adonis()' function [@Oksanen2016] to perform permutational multivariate analysis of variance to test whether treatment plots formed distinct groupings.
To test whether treatment plots were equally dispersed, or not, we used the 'vegan::betadisper' function [@Oksanen2016].

@@

# RESULTS

@@

# DISCUSSION

@@

# ACKNOWLEDGEMENTS

@@

# FUNDING

@@

# TABLES

@@

# FIGURES

@@

# REFERENCES

### 2017-07-11 manuscript/usses_water_ms.tex
trivia
more intro writing

\documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions

@@

\section{Introduction}\label{introduction}

@@

Manipulating potentially limiting resources offers a route to
understanding how ecosystems will respond to resource levels that fall
outside the historical range of variability \citep{Knapp2017}. Chronic
alterations to resource availability should cause community composition
to shift over time, the focal ecosystem is sensitive to the manipulated
resource. Species' relative abundances are expected to re-order and,
eventually, some species will be lost from the local community and new
species will be gained \citep{Smith2009, Avolio2015}. On the heels of
changes in community composition, subsequent changes in the level of
ecosystem functioning may occur.

\citet{Smith2009} introduced the `Hierarchical Response Framework' for
understanding the interplay of community composition and ecosystem
functioning in response to resource manipulations over time. In the near
term, ecosystem functioning such as annual net primary productivity
(ANPP) will reflect the physiological responses of individual species to
the manipulated resource level. For example, ANPP may decline under
simulated drought because the initial community consisted of
drought-intolerant species \citep{Hoover2014}. Over longer time spans,
ecosystem functioning may recover as new species colonize or initial
species reorder in relative abundance. For example, ANPP may initially
decline, but eventually rise back to pre-treatment levels once
drought-tolerant species become more abundant and compensate for
drought-intolerant species \citep{Hoover2014}. It is also possible that
ecosystem functioning shifts to a new mean state, reflective of the
suite of species in the new community \citep{Knapp2012}.

Here we report the results of a five-year precipitation manipulation
experiment in a sagebrush steppe grassland. We imposed drought and
irrigation treatments (50\% decrease/increase) and measured ecosystem
(ANPP) and community (species composition) responses. We focus on how
the drought and irrigation treatments affect the relationship between
ANPP and available soil moisture, and if community dynamics underly the
ecosystem responses. In particular, we are interested in the consistency
of the ANPP-soil moisture relationship among treatments. Is the
relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high
soil moisture? To answer these questions we fit a random intercepts,
random slopes model to test whether the regressions differed among
treatments.

\section{Materials and Methods}\label{materials-and-methods}

@@

To see if community composition differed among treatments through time,
we used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis
distances. For each year of the experiment, we first calculated
Bray-Curtis distances among all plots, and then extracted those
distances for use in the NMDS. Because we standardized species'
abundances, some values were negative, which is not allowed for
calculating Bray-Curtis distances. We simply added `2' to each abundance
value to ensure all values were greater than zero. We plotted the first
two axes of NMDS scores to see if community composition overlapped, or
not, among treatments in each year. We used the `metaMDS()' function in
the R package `vegan' \citep{Oksanen2016} to calculate Bray-Curtis
distances and then to run the NMDS analysis. We used the
`vegan::adonis()' function \citep{Oksanen2016} to perform permutational
multivariate analysis of variance to test whether treatment plots formed
distinct groupings. To test whether treatment plots were equally
dispersed, or not, we used the `vegan::betadisper' function
\citep{Oksanen2016}. Lastly, we examined rank clocks of species'
abundances through time to assess the stability of community composition
over the course of the experiment \citep{Collins2008}. Rank clocks were
made using `ggplot2' \citep{Wickham2009} and R code from the `codyn'
package \citep{Hallett2016}.

@@

\section{Results}\label{results}

@@

\section{Discussion}\label{discussion}

@@

\section{Acknowledgments}\label{acknowledgments}

@@

\section{Funding}\label{funding}

@@

\section{Tables}\label{tables}

@@

\section{Figures}\label{figures}

>>>>

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

@@

\section{INTRODUCTION}\label{introduction}

@@

Another problem with relying on historical ecoystem functional responses
is that they are not static. Changes in species identities and
abundances can alter an ecosystem's functional response to water
availability because different species have different physiological
thresholds for producting biomass. \citet{Smith2009} introduced the
`Hierarchical Response Framework' (HRF) for understanding the interplay
of community composition and ecosystem functioning in response to
resource manipulations over time. In the near term, ecosystem
functioning such as annual net primary productivity (ANPP) will reflect
the physiological responses of individual species to the manipulated
resource level. For example, ANPP may decline under simulated drought
because the initial community consisted of drought-intolerant species
\citep{Hoover2014}. Over longer time spans, ecosystem functioning may
recover as new species colonize or initial species reorder in relative
abundance. For example, ANPP may initially decline, but eventually rise
back to pre-treatment levels once drought-tolerant species become more
abundant and compensate for drought-intolerant species
\citep{Hoover2014}. It is also possible that ecosystem functioning
shifts to a new mean state, reflecting the suite of species in the new
community \citep{Knapp2012}.

Manipulating potentially limiting resources offers a route to
understanding how ecosystems will respond to resource levels that fall
outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. Following the HRF, we propose four alternative predictions
for the effect of precipitation manipulation on the ecosystem functional
response to soil moisture, that is, the soil moisture-ANPP relationship
(Figure 1). The four predictions are based on possible outcomes at the
community (e.g., community composition) and ecosystem (e.g., soil
moisture-ANPP regression) levels.

First, altered precipitation changes neither ecosystem functional
response nor community composition (Fig. 1, top left). In this case,
changes in ANPP simply follow the soil moisture-ANPP relationship under
ambient conditions. This corresponds to the early phases of the HRF,
where ecosystem response is due to the physiological responses of
individual species. Second, the ecosystem functional response changes
but community composition remains the same (Fig. 1, top right). A
saturating soil moisture-ANPP response fits this scenario, where
individual species hit physiological thresholds or are limited by some
other resource. Third, the ecosystem functional response is consistent
but underlying community composition changes (Fig. 1, bottom left). In
this case, changes in species' indentities or abundances occur in
response to altered precipitation levels and species more suited to the
new conditions compensate for reduced function of initial species.
Fourth, and last, both ecosystem functional response and community
composition change (Fig. 1, bottom right). New species, or newly
abundant species, with different physiological responses completely
reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. Thus, our task is not to test
the validity of the HRF, but rather to amass information on how quickly
ecosystem functional responses change in different ecosystems. Likewise,
we need to understand whether changes at the ecosystem level are driven
community level changes or individual level responses.

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (50\% decrease/increase) and measured
ecosystem (ANPP) and community (species composition) responses. We focus
on how the drought and irrigation treatments affect the relationship
between available soil moisture and ANPP, and if community dynamics
underly the ecosystem responses. In particular, we are interested in the
consistency of the soil moisture-ANPP relationship among treatments. Is
the relationship steeper under the drought treatment, at low soil
moisture? Does the relationship saturate under the irrigation treatment,
at high soil moisture? To answer these questions we fit a random
intercepts, random slopes model to test whether the regressions differed
among treatments. We also analyzed community composition over time,
allowing us to place our experimental results within the framework of
the HRF and our competing predictions (Fig. 1).

\section{METHODS}\label{methods}

@@

To see if community composition differed among treatments through time,
we used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis
distances. For each year of the experiment, we first calculated
Bray-Curtis distances among all plots, and then extracted those
distances for use in the NMDS. Because we standardized species'
abundances, some values were negative, which is not allowed for
calculating Bray-Curtis distances. We simply added `2' to each abundance
value to ensure all values were greater than zero. We plotted the first
two axes of NMDS scores to see if community composition overlapped, or
not, among treatments in each year. We used the `metaMDS()' function in
the R package `vegan' \citep{Oksanen2016} to calculate Bray-Curtis
distances and then to run the NMDS analysis. We used the
`vegan::adonis()' function \citep{Oksanen2016} to perform permutational
multivariate analysis of variance to test whether treatment plots formed
distinct groupings. To test whether treatment plots were equally
dispersed, or not, we used the `vegan::betadisper' function
\citep{Oksanen2016}.

@@

\section{RESULTS}\label{results}

@@

\section{DISCUSSION}\label{discussion}

@@

\section{ACKNOWLEDGEMENTS}\label{acknowledgements}

@@

\section{FUNDING}\label{funding}

@@

\section{TABLES}\label{tables}

@@

\section{FIGURES}\label{figures}
